Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-09-22 src/tex/ms.tex
[showyourwork] overleaf sync

    As a first application example for real data we choose a set of short exposure observations of the SNR \textit{1E~0102.2–7219} by the \chandra telescope and combine them into a single deep observation using \jolideco and the patch prior learned from JWST data. In total we chose 25 observations\footnote{Same list of observation IDs as for \url{https://chandra.si.edu/photo/2009/e0102/}} taken with  the ACIS instrument and no grating.

    We reduced the data using \texttt{ciao} v4.15 and a custom \texttt{snakemake} workflow (see also Section~\ref{sec:reproducibility} on reproducibility). For the counts image we selected the energy range from \qty[mode = text]{0.5}{keV} to  \qty[mode = text]{7}{keV} and a pixel size of $0.25$ of the native pixel size of \chandra. This corresponds to an absolute  pixel size of $\approx$\qty[mode = text]{0.125}{$\arcsec$/pixel} or a sampling of $\approx8$ pixels for the $38\%$ containment radius of Chandra's PSF. \chandra's instrument response varies with the observation offset and energy \todo{reference}, so we simulated the PSF for each observation independently using the tool \texttt{marx} \citep{Davis2012} and the \texttt{simulate\_psf} script. As reference spectrum for the PSF simulation we assumed the model derived by \cite{Plucinsky2017}. We converted the spectral from XSPEC to sherpa script using the helper script \texttt{convert\_xspec\_script}. As reported in \cite{Xi2019} the effect of pileup, even for the brightest regions is small ($<5\%$), so we did not include the effect of pileup in our PSF simulations.

    We assume a spatially constant background of \qty[mode = text]{1e-5}{cts/pixel} but allow the norm to vary later in the reconstruction process. We assume the exposure to be spatially constant and express its absolute value for each observation as the ratio of observation time with respect to a reference observation. As reference we chose the observation with id 1308. For \jolideco we used the GMM prior learned on the JWST Cas A image. As there is a lot of statistics in the image we additonally used an oversampling factor of 2. For each observation we allowed the background level and position calibration to vary. As reference for the positional shift we also used the observation with the ID 1308.

@@

@@

    \subsection{\fermi Event Types}
      \begin{figure*}
        \script{example-fermi-lat.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-fermi-lat.pdf}
            \caption{
                Illustration of the result of the \jolideco method applied to \fermi data of the supernova remnant \textit{RX J0852.0-4622} or \textit{Vela Junior}. The left image shows the counts above \qty[mode = text]{10}{GeV}. The different event classes are stacked into a single image. The image in the center shows the flux reconstructed by the \jolideco method. The image on the right shows the residuals as computed by Equation~\ref{eq:approx-sigma} and smoothed with a \textit{Gaussian} kernel of width 5~pixels. More information on this analysis example can be found on \url{https://github.com/jolideco/jolideco-fermi-examples}. \todo{Move color bar to top and make images larger}
            }
            \label{fig:example-fermi-lat}
        \end{centering}
    \end{figure*}
    As a second example we apply the \jolideco method to data from the Fermi Large Area Telescope (LAT). The LAT is a pair conversion \gammaray telescope on board of the Fermi satellite. It operates in the energy range between \qty[mode = text]{30}{MeV} to  \qty[mode = text]{300}{GeV}. For this example we chose again a Galactic source with complex morphology: the supernova remnant Vela Junior. We use 14 years (from 2008 August 4 to 2023 August 4) of \endquote{P8R3\_SOURCE} photons with reconstructed energy in the range of \qty[mode = text]{10}{GeV} to  \qty[mode = text]{2}{TeV} and applied a zenith angle cut of \ang{105}. To preserve as much information as possible on the angular resolution of the instrument we split the data into the four event types \enquote{PSF0}, \enquote{PSF1}, \enquote{PSF2} and \enquote{PSF3}, as defined by the \fermi collaboration. We chose a pixel size of \qty[mode = text]{0.02}{$\deg$}, which corresponds to approximately 1/5th of the size of the PSF in the best event class. For the first part of the data reduction we use the standard \texttt{fermitools v2.2.0} \cite{}.

@@

>>>>

    \subsubsection{Image Example}
    As a first application example for real data we choose a set of short exposure observations of the SNR \textit{1E~0102.2–7219} by the \chandra telescope and combine them into a single deep observation using \jolideco and the patch prior learned from JWST data. In total we chose 25 observations\footnote{Same list of observation IDs as for \url{https://chandra.si.edu/photo/2009/e0102/}} taken with  the ACIS instrument and no grating.

    We reduced the data using \texttt{ciao} v4.15 and a custom \texttt{snakemake} workflow (see also Section~\ref{sec:reproducibility} on reproducibility). For the counts image we selected the energy range from \qty[mode = text]{0.5}{keV} to  \qty[mode = text]{7}{keV} and a pixel size of $0.25$ of the native pixel size of \chandra. This corresponds to an absolute  pixel size of $\approx$\qty[mode = text]{0.125}{$\arcsec$/pixel} or a sampling of $\approx2$ pixels for the $38\%$ containment radius of Chandra's PSF. \chandra's instrument response varies with the observation offset and energy \todo{reference}, so we simulated the PSF for each observation independently using the tool \texttt{marx} \citep{Davis2012} and the \texttt{simulate\_psf} script. As reference spectrum for the PSF simulation we assumed the model derived by \cite{Plucinsky2017}. We converted the spectral from XSPEC to sherpa script using the helper script \texttt{convert\_xspec\_script}. As reported in \cite{Xi2019} the effect of pileup, even for the brightest regions is small ($<5\%$), so we did not include the effect of pileup in our PSF simulations.

    We assume a spatially constant background of \qty[mode = text]{1e-5}{cts/pixel} but allow the norm to vary later in the reconstruction process. We assume the exposure to be spatially constant and express its absolute value for each observation as the ratio of observation time with respect to a reference observation. As reference we chose the observation with id 1308. For \jolideco we used the GMM prior learned on the JWST Cas A image. As there is a lot of statistics in the image we additionally used an oversampling factor of 2. For each observation we allowed the background level and position calibration to vary. As reference for the positional shift we also used the observation with the ID 1308.

@@

    \newpage
    \subsubsection{Flux Profile}
    \begin{figure*}
        \script{chandra-e0102-zoom-a.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/chandra-e0102-zoom-a.pdf}
            \caption{
               Flux and counts profile of the Zoom A example region. The associated $2\sigma$ error band estimated using the bootstrap method is shown transparent in the background.
            }
            \label{fig:chandra-e0102-zoom-a}
        \end{centering}
    \end{figure*}
    To get a more qualitative measurement of the improvement we compute a flux profile on the image reconstructed by \jolideco. For this we chose a region perpendicular to the filament shown in Zoom A of Figure~\ref{fig:example-chandra}. To estimate the errors we use the bootstrapping method. For each individual observation we sample from the event list with replacement. For each realization of the set of observation we run \jolideco and compute a flux profile on the reconstructed image. This way we obtain a distribution of profiles, which we can use to compute an error band. The result is shown in Figure~\ref{fig:chandra-e0102-zoom-a}.

    We find that the error is

    We approximately expect the following relation between the width of the counts profile, the PSF and the flux profile:

    \begin{equation}
        \sigma_{\mathrm{Flux}} = \sqrt{\sigma_{\mathrm{Counts}}^2 - \sigma_{\mathrm{PSF}} ^2}
    \end{equation}

    With $\sigma_{\mathrm{PSF}} \approx 0.25 \mathrm{arcsec}$ we obtain ..., which is consistent with the values we measured for the profile.

@@

    \subsection{\fermi Event Types}
    As a second example we apply the \jolideco method to data from the Fermi Large Area Telescope (LAT). The LAT is a pair conversion \gammaray telescope on board of the Fermi satellite. It operates in the energy range between \qty[mode = text]{30}{MeV} to  \qty[mode = text]{300}{GeV}. For this example we chose again a Galactic source with complex morphology: the supernova remnant Vela Junior. We use 14 years (from 2008 August 4 to 2023 August 4) of \endquote{P8R3\_SOURCE} photons with reconstructed energy in the range of \qty[mode = text]{10}{GeV} to  \qty[mode = text]{2}{TeV} and applied a zenith angle cut of \ang{105}. To preserve as much information as possible on the angular resolution of the instrument we split the data into the four event types \enquote{PSF0}, \enquote{PSF1}, \enquote{PSF2} and \enquote{PSF3}, as defined by the \fermi collaboration. We chose a pixel size of \qty[mode = text]{0.02}{$\deg$}, which corresponds to approximately 1/5th of the size of the PSF in the best event class. For the first part of the data reduction we use the standard \texttt{fermitools v2.2.0} \cite{}.

@@

      \begin{figure*}
        \script{example-fermi-lat.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-fermi-lat.pdf}
            \caption{
                Illustration of the result of the \jolideco method applied to \fermi data of the supernova remnant \textit{RX J0852.0-4622} or \textit{Vela Junior}. The left image shows the counts above \qty[mode = text]{10}{GeV}. The different event classes are stacked into a single image. The image in the center shows the flux reconstructed by the \jolideco method. The image on the right shows the residuals as computed by Equation~\ref{eq:approx-sigma} and smoothed with a \textit{Gaussian} kernel of width 5~pixels. More information on this analysis example can be found on \url{https://github.com/jolideco/jolideco-fermi-examples}. \todo{Move color bar to top and make images larger}
            }
            \label{fig:example-fermi-lat}
        \end{centering}
    \end{figure*}
